What does the crowd believe? A hierarchical approach to estimating subjective beliefs from empirical data

### 2016-02-01 text/01_CogSci_abstract/paper.tex
edits throughout methods and discussion

\author{{\large \bf Michael Franke} \\
  michael.franke@uni-tuebingen.de \\
  Department of Linguistics, Wilhelmstra\ss e 19 \\
 72074 T\"{u}bingen, Germany \AND 
 {\large \bf Erin Bennett, Judith Degen, Michael Henry Tessler, Noah D. Goodman}\\
     \{ebennett,jdegen,mhtessler,ngoodman\}@stanford.edu \\
     Department of Psychology, 450 Serra Mall \\
  Stanford, CA 94305 USA }

@@

The data we would like to explain are: (i) the normalized slider ratings $s_{ijk} \in [0;1]$ of
participant $i \in \set{1, \dots, 20}$ for item $j \in \set{1, \dots, 8}$ and bin
$k \in \set{1, \dots, 15}$ from the binned histogram task; (ii) the bins $n_{ij} \in \set{1, \dots, 15}$ in which participant
$i$'s number choice for item $i$ was in the give-a-number task; and (iii) the binary choices $c_{ijl} \in \set{0,1}$ of
whether participant $i$ selected the higher bin for item $j$ in the paired comparison task for each comparison
$l$. There are two simplifications in need of commenting. In (i), we focus on slider ratings after
normalizing for each participant, because we assume that slider adjustments reflect relative, not
absolute estimates of subjective beliefs. In (ii), we focus on bin choices, not actual number
choices, in order to avoid, as much as possible, considerations of salience of particular
numbers, and also because otherwise data from items with smaller domains of plausible numbers
would get more weight than data from items which allow for a wider set of number choices.

@@

The link function for the slider rating data uses a logit transformation to project observed
slider ratings $s_{ijk}$ and latent probabilities $P_{ijk}$, which are bound to lie between 0
and 1, to the reals. The likelihood of logit-transformed observation $s_{ijk}$ is given by a
Gaussian with standard deviation $\sigma$ around the logit-transformed predictor $P_{ijk}$. On
top of that, there is a parameter $\kappa$, the steepness of the logit transform of $P_{ijk}$,
that allows response likelihoods to capture end-point affinity for $\kappa >1$ (values of
$P_{ijk}$ close to 0 or 1 are likely mapped to 0 or 1) or end-point aversion for $\kappa <0$
(values of $P_{ijk}$ are likely to be realized as more median), with a prior that expects
$\kappa=1$.
\begin{align*}
  \text{logit}(s_{ijk}) &\sim
        \text{Norm}(\text{logit}(P_{ijk}, \kappa), \sigma) \\
        \sigma &  \sim \text{Gamma}(0.0001,0.0001) & \kappa &\sim \text{Gamma}(5,5)
\end{align*}

The link function for number choice data treats each bin $n_{ij}$ as a draw from a categorical
distribution where the probability of bin $k$ is proportional to $\expo(a P_{ij})$, i.e., a
soft-max choice from $P_{ij}$. The higher parameter $a$, the more likely $n_{ij}$ is the mode
of $P_{ij}$. For $a \rightarrow 0$, all bins become equiprobable.
\begin{align*}
  n_{ij} & \sim
        \text{Categorical}(\expo(a P_{ij})) &
 a & \sim \text{Gamma}(2,1)
\end{align*}

Finally, consider the link function for bin comparisons. We are interested in the likelihood
with which participant $i$ selects the higher bin for item $j$ in bin comparison condition
$l$. Suppose $l$ is about comparing the lower bin $b_l$ to the higher $b_h$. The perhaps most
natural approach would be to link the likelihood of choosing $b_h$ over $b_l$ to the difference
between $P_{ijb_h}$ and $P_{ijb_l}$. However, as discussed above, this does not appear to be
what participants were doing. Indeed, a model that implements this idea failed to capture the
relevant regularities in the data blatantly. Another plausible link function is to assume that
what matters in bin comparisons is the distance to the mode of $P_{ij}$: soft-max prefer the
bin that is closer to the mode of $P_{ij}$; select randomly if both bins are equally far from
the prototype.\footnote{Notice that
  $(1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} = \frac{\exp(b p^{\text{high}})}{\exp(b
    p^{\text{high}}) + \exp(b p^{\text{low}})}$
  with $p^\text{low}_{ijl} = 2 - p^\text{high}_{ijl}$.}
\begin{align*}
  c_{ijl} & \sim \text{Bern}( (1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} ) \ \ \ 
  b  \sim \text{Gamma}(2,1) \\
  p^\text{high}_{ijl} & = \begin{cases}
    2 & \text{if $mode(P_{ij})$ is closer to higher} \\
    & \text{ bin of $l$ than to lower bin } \\ 1 & \text{if equal
      distance} \\ 0 & \text{otherwise}  
  \end{cases}
\end{align*}

@@

Of most theoretical interest are posterior estimates of $Q_j$. \figref{fig:PosteriorQj}
shows means of the posterior $Q_j$, with their 95\% HDIs, alongside the averaged normalized
slider ratings. The latter provide a very good approximation of the latent, inferred
population-level beliefs. Inspection of posteriors of individual $P_{ij}$ shows that there is
ample variation between participants. Nonetheless, the way the model suggest we should think about
harnessing the individual $P_{ij}$s under a population-level central tendency is closely
approximated by simply averaging over slider ratings. Although the match is clearly not
perfect, it is good enough to maintain that the latter are a reasonable (enough) and practical
way of assessing what the crowd believes despite individual differences.

@@

Inferences based on the model are only as reliable as the model itself is plausible. Model
criticism is therefore important. \figref{fig:PPCs} shows posterior predictive checks at
the population-aggregate level for all of our three task types. For the binned histogram task,
posterior predictions are spot-on. Some of the give-a-number data is surprising even for
the model trained on this very data. This could have various reasons: (i) the give-a-number data does
not have a huge influence on the posterior likelihood, (ii) number choices may be influenced by
saliency and/or roundness of numbers after all. Finally, there is one condition in the paired
comparison task that the model definitely got wrong. This is the choice of what is more likely:
that one or that none of 14 marbles thrown into a pool would sink. The model would predict that
almost everybody should answer that it is more likely that one marble sank. But that is not
what we observe. It may be that participants revise beliefs about ``normality'' of the marbles,
while holding on to an assumption that all marbles behave in the same way 
\cite{DegenTessler2015:Wonky-worlds:-L}.

@@

Posterior predictive checks indicate that the trained model captures patterns of answers at the
aggregate population level well. To have a more fine-grained measure of model fit, we also
looked at posterior predictive $p$-values \citep{GelmanCarlin2014:Bayesian-Data-A} at the level
of participants and items. We focus on the binned histograms task, because here the
population-level posterior predictive checks revealed no systematic deviance. For the slider
rating task, fixing a participant and an item, observations and replicates are probability
vectors of length 15. In a first analysis, we used the mean of these probability vectors as a
test statistic. The minimum posterior predictive $p$-value over all 20 (participants) times 8
(items) cases was 0.13, suggesting that the means of observed $s_{ij}$ are non-surprising to
the trained model. In a second analysis, we used entropy as a test statistic. Two cases gave
posterior predictive $p$-values lower than 0.05. These were from the two participants who gave
a very extreme slider rating for the ``marbles'' item, basically assigning all ``mass'' to the
last bin. What this suggests is that the model can cope reasonably well also with
individual-level data, but has problems accounting for ``extreme'' choices, given that the
population-level hyper-prior on $P_{ij}$ will lead to shrinkage.

\section{Conclusion}

>>>>

\author{{\large \bf Michael Franke} \\
  michael.franke@uni-tuebingen.de \\
  Department of Linguistics, Wilhelmstra\ss e 19 \\
 72074 T\"{u}bingen, Germany \AND 
 {\large \bf Erin Bennett, Judith Degen, Michael Henry Tessler, Noah D. Goodman}\\
     \{ebennett, jdegen, mtessler, ngoodman\}@stanford.edu \\
     Department of Psychology, Stanford University \\
     450 Serra Mall, Stanford, CA 94305 USA }

@@

The data we would like to explain are: (i) the normalized slider ratings $s_{ijk} \in [0;1]$ of
participant $i \in \set{1, \dots, 20}$ for item $j \in \set{1, \dots, 8}$ and bin
$k \in \set{1, \dots, 15}$ from the binned histogram task; (ii) the bins $n_{ij} \in \set{1, \dots, 15}$ in which participant
$i$'s number choice for item $j$ was in the give-a-number task; and (iii) the binary choices $c_{ijl} \in \set{0,1}$ of
whether participant $i$ selected the higher bin for item $j$ in the paired comparison task for each comparison
$l$. There are two simplifications in need of commenting. In (i), we focus on slider ratings after
normalizing for each participant, because we assume that slider adjustments reflect relative, not
absolute estimates of subjective beliefs. In (ii), we focus on bin choices, not actual number
choices, in order to avoid, as much as possible, considerations of salience of particular
numbers, and also because otherwise data from items with smaller domains of plausible numbers
would get more weight than data from items which allow for a wider set of number choices.

@@

The link function for the \textbf{slider rating data} uses a logit transformation to project observed
slider ratings $s_{ijk}$ and latent probabilities $P_{ijk}$, which are bound to lie between 0
and 1, to the reals. The likelihood of logit-transformed observation $s_{ijk}$ is given by a
Gaussian with standard deviation $\sigma$ around the logit-transformed predictor $P_{ijk}$. On
top of that, there is a parameter $\kappa$, the steepness of the logit transform of $P_{ijk}$,
that allows response likelihoods to capture end-point affinity for $\kappa >1$ (values of
$P_{ijk}$ close to 0 or 1 are likely mapped to 0 or 1) or end-point aversion for $\kappa <0$
(values of $P_{ijk}$ are likely to be realized as more median), with a prior that expects
$\kappa=1$.
\begin{align*}
  \text{logit}(s_{ijk}) &\sim
        \text{Norm}(\text{logit}(P_{ijk}, \kappa), \sigma) \\
        \sigma &  \sim \text{Gamma}(0.0001,0.0001) & \kappa &\sim \text{Gamma}(5,5)
\end{align*}

The link function for \textbf{number choice data} treats each bin $n_{ij}$ as a draw from a categorical
distribution where the probability of bin $k$ is proportional to $\expo(a P_{ij})$, i.e., a
soft-max choice from $P_{ij}$. The higher parameter $a$, the more likely $n_{ij}$ is the mode
of $P_{ij}$. For $a \rightarrow 0$, all bins become equiprobable.
\begin{align*}
  n_{ij} & \sim
        \text{Categorical}(\expo(a P_{ij})) &
 a & \sim \text{Gamma}(2,1)
\end{align*}

Finally, consider the link function for \textbf{bin comparisons}. We are interested in the likelihood
with which participant $i$ selects the higher bin for item $j$ in bin comparison condition
$l$. Suppose $l$ is about comparing the lower bin $b_l$ to the higher $b_h$. The perhaps most
natural approach would be to link the likelihood of choosing $b_h$ over $b_l$ to the difference
between $P_{ijb_h}$ and $P_{ijb_l}$. However, as discussed above, this does not appear to be
what participants were doing. Indeed, a model that implements this idea blatantly failed to capture the
relevant regularities in the data. Another plausible link function is to assume that
what matters in bin comparisons is the distance to the mode of $P_{ij}$: soft-max prefer the
bin that is closer to the mode of $P_{ij}$; select randomly if both bins are equally far from
the prototype.\footnote{Notice that
  $(1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} = \frac{\exp(b p^{\text{high}})}{\exp(b
    p^{\text{high}}) + \exp(b p^{\text{low}})}$
  with $p^\text{low}_{ijl} = 2 - p^\text{high}_{ijl}$.
  \mht{I am unable to notice this. Double check? Intuition tells me $p^\text{low}_{ijl} = 1 - p^\text{high}_{ijl}$.}
  }
\begin{align*}
  c_{ijl} & \sim \text{Bern}( (1 + \exp(2b(1-p^{\text{high}}_{ijl})) )^{-1} ) \ \ \ 
  b  \sim \text{Gamma}(2,1) \\
  p^\text{high}_{ijl} & = \begin{cases}
    2 & \text{if $mode(P_{ij})$ is closer to higher} \\
    & \text{ bin of $l$ than to lower bin } \\ 1 & \text{if equal
      distance} \\ 0 & \text{otherwise}  
  \end{cases}
\end{align*}

@@

Of most theoretical interest are posterior estimates of $Q_j$. \figref{fig:PosteriorQj}
shows means of the posterior $Q_j$, with their 95\% HDIs, alongside the averaged normalized
slider ratings. The latter provide a very good approximation of the latent, inferred
population-level beliefs. Inspection of posteriors of individual $P_{ij}$ shows that there is
ample variation between participants. Nonetheless, the way the model suggests we should think about
harnessing the individual $P_{ij}$s under a population-level central tendency is closely
approximated by simply averaging over slider ratings. Although the match is clearly not
perfect, it is good enough to maintain that the latter are a reasonable (enough) and practical
way of assessing what the crowd believes despite individual differences.

@@

Inferences based on the model are only as reliable as the model itself is plausible. Model
criticism is therefore important. \figref{fig:PPCs} shows posterior predictive checks at
the population-aggregate level for all of our three task types. For the binned histogram task,
posterior predictions are spot-on. For the give-a-number task, some of the data is surprising despite 
the model being trained on this very data. This could have various reasons: (i) the give-a-number data does
not have a huge influence on the posterior likelihood, (ii) number choices may be influenced by
saliency and/or roundness of numbers after all (and thus, not accurately reflect the true beliefs $Q_j$). Finally, there is one condition in the paired
comparison task that the model definitely got wrong. This is the choice of what is more likely:
One or none of 14 marbles thrown into a pool would sink. The model would predict that
almost everybody should answer that it is more likely that one marble sank. But that is not
what we observe. It may be that participants revise beliefs about ``normality'' of the marbles,
while holding on to an assumption that all marbles behave in the same way 
\cite{DegenTessler2015:Wonky-worlds:-L}.

@@

Posterior predictive checks indicate that the trained model captures patterns of answers at the
aggregate population level well. To have a more fine-grained measure of model fit, we also
looked at posterior predictive $p$-values \citep{GelmanCarlin2014:Bayesian-Data-A} at the level
of participants and items. We focus on the binned histograms task, because here the
population-level posterior predictive checks revealed no systematic deviance. For the slider
rating task, fixing a participant and an item, observations and replicates are probability
vectors of length 15. In a first analysis, we used the mean of these probability vectors as a
test statistic. The minimum posterior predictive $p$-value over all 20 (participants) times 8
(items) cases was 0.13, suggesting that the means of observed $s_{ij}$ are non-surprising to
the trained model. In a second analysis, we used entropy as a test statistic. Two cases gave
posterior predictive $p$-values lower than 0.05. These were from the two participants who gave
a very extreme slider rating for the ``marbles'' item, basically assigning all ``mass'' to the
last bin. What this suggests is that the model can cope reasonably well also with
individual-level data, but has problems accounting for ``extreme'' choices, given that the
population-level hyper-prior on $P_{ij}$ will lead to shrinkage.
\mht{Is it worth mentioning about the possilbiity of including subject-specific link function parameters e.g., $\kappa$ and $\alpha$?}
\section{Conclusion}
